---
title: "Homework_12"
author: "Luke Daniels"
date: "4/11/2018"
output: html_document
---

For this ggplot demonstration, I will be using the R data set "USarrest".  Here is the structure: 

```{r}
library(ggplot2)
library(patchwork)
library(maps) 
library(readxl)
library(openxlsx)
head(USArrests)
```

As we can see this is state / location data.  It would be nice to view this data on a map to observe general trends. For this we utilize the "maps" package. 

```{r}
crimes <- data.frame(state = tolower(rownames(USArrests)), USArrests)



states_map <- map_data("state")
map1 <- ggplot(crimes, aes(map_id = state)) + geom_map(aes(fill = Murder), map = states_map) + expand_limits(x = states_map$long, y = states_map$lat)
last_plot() + coord_map()


states_map2 <- map_data("state")
map2 <- ggplot(crimes, aes(map_id = state)) + geom_map(aes(fill = Assault), map = states_map2) + expand_limits(x = states_map$long, y = states_map$lat)
last_plot() + coord_map()



```


To see the initial trend we will plot geom_smooth using 'loess' 

```{r}
Plot1 <- ggplot(USArrests, mapping = aes(x = UrbanPop, y = Assault)) + 
  geom_point()+
  geom_smooth()
Plot1
```

\ 


We can view the relationship between Population and Assualt Rates for each state.  However I acknowledge that this is a **bad** method since we cannot pinpoint each state.  It does demonstrate the power of ggplot! 
```{r}
Data <- read_excel("Data.xlsx")

gg <- ggplot(Data, aes(x=UrbanPop, y=Assault)) + 
  geom_point(aes(col=State)) + 
  geom_smooth(method="loess", se=F) + 
  labs(subtitle="Population Vs Assault", 
       y="Assalt", 
       x="UrbanPopulation", 
       title="Scatterplot")

plot(gg)
```

Furthermore we could make each data point a function of Rape levels as well.  To do this we map the points to size = Rape. 

```{r}
g3 <- ggplot(Data, aes(x=UrbanPop, y=Assault)) + 
  geom_point(aes(col=State, size = Rape)) + 
  geom_smooth(method="loess", se=F) + 
  labs(subtitle="Population Vs Assault", 
       y="Assalt", 
       x="UrbanPopulation", 
       title="Scatterplot")

plot(g3)
```
As we discussed in class, this is information overload!  But we have managed to represent four different variables: State, Population, Assault and Rape. 

\ 

Maybe a simplier way to view these relationships is by using text to represent each point.  We can easily change the test size as well.  The command "check_overlap = TRUE" is useful to make sure the text does not over lap so it is clearly visible! 
```{r}
g4 <- ggplot(Data,
       aes(UrbanPop, Murder, label = as.character(State))) + theme_classic() +
  geom_text(check_overlap = TRUE, size = 3) 
  
g4
```
\ 

These types of graphs can be manipulated very easily.  If we ever wanted to angle the text we simply indicate the angle we want.  In this case its a 45 degree angle. 

```{r}
g5 <- ggplot(Data,
             aes(UrbanPop, Murder, label = as.character(State))) + theme_classic() +
  geom_text(angle =45, size = 3)
g5
```


\ 

Using 'Patchwork' we can combine multiple graphs 
```{r}
map1 + Plot1 - g5 + plot_layout(ncol = 1)
```

