---
title: " Probability Distributions in R"
author: "Luke Daniels"
date: "2/21/2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *Loading the Data*
I used a random data file for ease of use.  

```{r}
library(MASS)
library(ggplot2)
library(haven) 
driving <- read_dta("driving_2004-2.dta")

```

# *Plotting the Initial Histogram*

```{r}
driving2 <- data.frame(driving$totfat)
                       names(driving2) <- c("totalfatal")
      
mean(driving2) 

p1 <- ggplot(data=driving2, aes(x=totalfatal, y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) 
print(p1)
```


# **Adding the Empirical Density Curve**
 
```{r}

p1 <-  p1 +  geom_density(linetype="dotted",size=0.75)
print(p1)

```

# **Get Maximum Likelihood Parameters for Normal** 

```{r}
normPars <- fitdistr(driving$totfat,"normal")
print(normPars)

normPars$estimate["mean"]

str(normPars)
```
# **Generating Different Distributions**

## **Normal Distribution**

```{r}
meanML <- normPars$estimate["mean"]
sdML <- normPars$estimate["sd"]

xval <- seq(0,max(driving$totfat),len=length(driving$totfat))

stat <- stat_function(aes(x = xval, y = ..y..), fun = dnorm, colour="red", n = length(driving$totfat), args = list(mean = meanML, sd = sdML))
print(p1 + stat)
```


## **Plot Exponential Probability Distribution**

 
```{r}
expoPars <- fitdistr(driving$totfat,"exponential")
rateML <- expoPars$estimate["rate"]

stat2 <- stat_function(aes(x = xval, y = ..y..), fun = dexp, colour="blue", n = length(driving$totfat), args = list(rate=rateML))
print(p1 + stat + stat2)
```


## **Uniform Probability Distribution** 

```{r}
stat3 <- stat_function(aes(x = xval, y = ..y..), fun = dunif, colour="darkgreen", n = length(driving$totfat), args = list(min=min(driving$totfat), max=max(driving$totfat)))
p1 + stat + stat2 + stat3
```

## **Gamma Distribution**

```{r}
gammaPars <- fitdistr(driving$totfat,"gamma")
shapeML <- gammaPars$estimate["shape"]
rateML <- gammaPars$estimate["rate"]

stat4 <- stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", n = length(driving$totfat), args = list(shape=shapeML, rate=rateML))
p1 + stat + stat2 + stat3 + stat4
```


## **Beta Distribution** 

```{r}
pSpecial <- ggplot(data=driving2, aes(x=driving$totfat/(max(driving$totfat + 0.1)), y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) + 
  xlim(c(0,1)) +
  geom_density(size=0.75,linetype="dotted")

betaPars <- fitdistr(x=driving$totfat/max(driving$totfat + 0.1),start=list(shape1=1,shape2=2),"beta")
shape1ML <- betaPars$estimate["shape1"]
shape2ML <- betaPars$estimate["shape2"]

statSpecial <- stat_function(aes(x = xval, y = ..y..), fun = dbeta, colour="orchid", n = length(driving$totfat), args = list(shape1=shape1ML,shape2=shape2ML))
pSpecial + statSpecial
```

## **Compare Log Likelihoods**

We want to find which distribution best fits the data.  The higher the number the better the fit. 

```{r}
normPars$loglik 
gammaPars$loglik 
expoPars$loglik

```

Gamma is the Best Fit! 

# **Create Simulated Data**

Creating simulated data based on the parameters above. 

```{r}
z <- rgamma(n= 48, shape=shapeML, rate=rateML) 
z <- data.frame(1:48,z)

names(z) <- list("ID","myVar")
z <- z[z$myVar>0,]
str(z)

p5 <- ggplot(data=z, aes(x=myVar, y=..density..)) +
  geom_histogram(color="grey60",fill="cornsilk",size=0.2) 
print(p5)

normPars <- fitdistr(z$myVar,"normal")
print(normPars)

str(normPars)

gammaPars <- fitdistr(z$myVar,"gamma")
shapeML <- gammaPars$estimate["shape"]
rateML <- gammaPars$estimate["rate"]

stat4 <- stat_function(aes(x = xval, y = ..y..), fun = dgamma, colour="brown", n = length(z$myVar), args = list(shape=shapeML, rate=rateML))
p5 + stat4

p1 + stat4

```

We can see that the simulated data given the parameters doesn't exactly match our original data set.  However, it is fairly good for simulated data! 
